---
title: "Logistic Regression and Random Forest to Determine Credit Card Default"
author: "Travis Gubbe"
date: "June 27, 2026"
output:
  html_document: default
  pdf_document: default
---

In this session, I examine the Credit Card Clients data set found on the UCI Machine Learning Repository website to determine if a person will default on their credit card using logistic regression and random forest. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(MASS)
library(caret)
library(neuralnet)
library(randomForest)
library(pander)
library(dplyr)
library(ggplot2)
library(glmnet)
library(ROSE)
library(Metrics)
library(car)
library(pROC) #ROC curve
library(OptimalCutpoints) #Optimal cuts for ROC curve
library(RColorBrewer) #R color palettes
library(glmnet)
```

## About the Data Set

The data set can be found on the UCI Machine Learning Repository site at the following link: <https://archive.ics.uci.edu/ml/datasets/default+of+credit+card+clients>.

The data contains 24 variables and a total of 30,000 individual instances. The variables are:

* Default Payment (0 = No, 1 = Yes)
* Amount of Given Credit
* Gender (1 = Male, 2 = Female)
* Education (1 = Graduate School, 2 = University, 3 = High School, 4 = Other, 5 = Unknown)
* Marital Status (1 = Married, 2 = Single, 3 = Others)
* Age
* History of Past Payment from April 2005 to September 2005 where -2=no consumption, -1=pay duly, 0=the use of revolving credit, 1=payment delay for one month, 2=payment delay for two months, … 9=payment delay for nine months and above
* Amount of Bill Statement from April 2005 to September 2005 
* Amount of Previous Payment April 2005 to September 2005

There is a separate variable for each past payment, bill statement, and previous payment from April to September.

## Exploratory Analysis

First, the data set is loaded into a data frame named "credit". The data frame is also viewed to see the columns and class types.

```{r exploratory}
data <- read.csv("~/R datasets/Credit_Card.csv")
credit = subset(data, select = c("Limit_Bal", "Sex", "Education", "Marriage", "Age", "Pay_Sep",
                                 "Pay_Aug", "Pay_July", "Pay_June", "Pay_May", "Pay_April",
                                 "Bill_Amt_Sep", "Bill_Amt_Aug", "Bill_Amt_July",
                                 "Bill_Amt_June", "Bill_Amt_May", "Bill_Amt_April",
                                 "Pay_Amt_Sep", "Pay_Amt_Aug", "Pay_Amt_July","Pay_Amt_June",
                                 "Pay_Amt_May", "Pay_Amt_April", "Default"))
#Inspect the data frame
head(credit)
#Inspect the classes of the data frame
sapply(credit, class)

attach(credit)
```

Next, I want to see the amount of missing values and duplicates in the data frame.

```{r cleaning}
sum(is.na(credit))

duplicates <- credit%>%duplicated()
duplicates_amount <- duplicates%>%(table)
duplicates_amount
```

Since there are 35 duplicates in the data, the data frame is filtered to remove the duplicates.

```{r distinct_values}
credit <- credit%>%distinct()
#Displays how many duplicates are present in the updated data frame.
duplicates_counts_unique <- credit%>%duplicated()%>%table()
duplicates_counts_unique
```

Next, the factor variables are converted from their numeric values to their actual names. This is done on a copy of the credit data frame. 

```{r copy_df}
credit1 <- data.frame(credit)
head(credit1)
```

```{r rename_variables}
#Rename factor variables to their appropriate settings
credit1$Sex[credit$Sex %in% "1"] = "Male"
credit1$Sex[credit$Sex %in% "2"] = "Female"

credit1$Education[credit$Education %in% "1"] = "Grad School"
credit1$Education[credit$Education %in% "2"] = "College"
credit1$Education[credit$Education %in% "3"] = "High School"
credit1$Education[credit$Education %in% "4"] = "Other"
credit1$Education[credit$Education %in% "5"] = "Unknown"

credit1$Marriage[credit$Marriage %in% "0"] = "Unknown"
credit1$Marriage[credit$Marriage %in% "1"] = "Married"
credit1$Marriage[credit$Marriage %in% "2"] = "Single"
credit1$Marriage[credit$Marriage %in% "3"] = "Other"

credit1$Default[credit$Default %in% "0"] = "No"
credit1$Default[credit$Default %in% "1"] = "Yes"
#See the change in the variable names
head(credit1)
```

Next, exploratory tables are made to view the distribution of the data set. 

### Exploratory Analysis

Next, bar plots and distribution tables are created to see the proportion of the variables. This is done to see if the data is normally distributed. If the data is not normally distributed, it's advantageous to see how the data is skewed.

```{r exploratory_tables}
#View the bar plots for the amount for each categorical variable
counts_Sex <- table(credit1$Sex)
barplot(counts_Sex, col = c("royalblue", "darkorange1"))
#Basic table view of the amount of males and females
table(credit1$Sex)
#Proportion of each gender in table
prop.table(counts_Sex)

counts_Education <- table(credit1$Education)
barplot(counts_Education, col = c("brown4", "green3", "mediumpurple2", "slategray3",
                                  "darkgoldenrod2"))
table(credit1$Education)
#Proportion of each education level in table
prop.table(counts_Education)

#Age groups divided into decades, then proportion table of each age group.
credit1$Age_Group <- cut(credit$Age,
                        breaks = c(20, 30, 40, 50, 60, 70, 80),
                        labels = c("20s", "30s", "40s", "50s", "60s", "70+"),
                        right = FALSE)

counts_Age <- table(credit1$Age_Group)
# Check distribution
prop.table(counts_Age)
#Bar plot of age groups with custom colors
colors <- brewer.pal(6, "Set3")
barplot(counts_Age, col = colors)
#Counts for each marriage status.
counts_Marriage <- table(credit1$Marriage)
barplot(counts_Marriage, col = c("magenta2", "Cyan3", "goldenrod"))
table(credit$Marriage)
#Proportion of each marriage status in table
prop.table(counts_Marriage)

counts_Default <- table(credit1$Default)
barplot(counts_Default, col = c("turquoise2", "sienna1"))
table(credit$Default)
prop.table(counts_Default)

table.default_gender <- table(credit1$Default, credit$Sex)
prop.table(table.default_gender, 2)
prop.table(table.default_gender, 1)
barplot(table.default_gender, col = c("sienna1", "royalblue"), beside = T,
        names.arg = c("Female", "Male"))
legend("topright", legend = c("No", "Yes"), fill = c("sienna1", "royalblue"))

ggplot(data = credit, aes(x = Age)) + geom_histogram(fill = "Blue", col = "Grey", bins = 30)

ggplot(data = credit, aes(x = Age)) + geom_histogram(aes(y = ..density..), fill = "Blue", col = "Grey", binwidth = 5)+geom_density(alpha = 0.2, color = "black", fill = "blue")
mean(credit1$Age)
```

Looking at the created charts and tables, the data has more females than males. In addition, the age group distribution is skewed to the right, meaning the data is represented by younger participants.

Another method for exploratory analysis is to investigate the correlation of the numeric variables. The correlation of numeric variables can help with model selection later on, as high correlation can indicate multicollinearity among variables, which can make it difficult for a model to separate and distinguish the importance of individual variables in the final model selection.

```{r correlation}
eda_numeric_only <- credit1[, c("Limit_Bal", "Age", 
                                "Bill_Amt_Sep", "Bill_Amt_Aug", "Bill_Amt_July", "Bill_Amt_June", "Bill_Amt_May",
                                "Bill_Amt_April", "Pay_Amt_Sep", "Pay_Amt_Aug", "Pay_Amt_July", "Pay_Amt_June",
                                "Pay_Amt_May", "Pay_Amt_April")]

# 2. Compute the Pearson correlation matrix safely
clean_eda_cor <- cor(eda_numeric_only, use = "complete.obs")

# 3. Print the perfectly aligned results rounded to 2 decimal places
print(round(clean_eda_cor, 2))
```
From the results, there are a few instances of high correlation between various numeric variables, particularly among variables with similar measurements (Bill_Amt_April and Bill_Amt_May). This will play a factor later in the analysis, particularly for the logistic regression model's regularization techniques.

### Train and Test Sets

Before creating prediction models, training and testing data sets are created. A training data set is a subset of examples used to train the model, while the testing data set is a subset used to test the training model.

```{r test_set_credit}
#Initializes number generator.
set.seed(123)
#New sample created for the training and testing data sets. The data is split with 75% in training and 25% in testing.
sample_split <- sample(c(TRUE, FALSE), nrow(credit), replace = TRUE, prob = c(0.75, 0.25))
train_set <- credit[sample_split, ]
test_set <- credit[!sample_split, ]

X_train <- model.matrix(Default ~ ., data = train_set)[, -1]
y_train <- as.numeric(train_set$Default)

# 2. Prepare the TESTING data matrices 
X_test  <- model.matrix(Default ~ ., data = test_set)[, -1]
y_test  <- as.numeric(test_set$Default)

```

## Logistic Regression

First, logistic regression is done to find the probability of default for an individual. Logistic regression models the probability that a response variable (Y) belongs to a particular category. This method uses maximum likelihood to fit the model in the range between 0 and 1. 

Logistic regression is a classification method great for a yes/no response. A number closer to 1 represents "Yes", while a number closer to 0 represents "No".

Before the creation of the model, An elastic net logistic regression model is first run as a means of feature selection and to prevent overfitting in the model. Elastic net is a hybrid version of the LASSO and ridge regularization methods, performing both coefficient shrinkage and feature selection which is appropriate in this case due to the high correlation among multiple variables. The elastic net has more flexibility than the LASSO regularization and is more robust than the Ridge regularization, selecting and dropping entire an entire group of correlated groups.

From the glmnet package, any alpha value between 0 and 1 is classified as elastic net. A for loop was created to determine the best alpha value among  when separating by between three alpha values (0.2, 0.5, and 0.8). 

```{r elastic_net, warning = FALSE}
library(glmnet)
#Elastic net for the data set
#Determine best alpha value
alphas <- c(0.2, 0.5, 0.8)
for (a in alphas) {
  set.seed(123)
  cv_model <- cv.glmnet(X_train, y_train, family = "binomial", alpha = a)
  cat("Alpha:", a, " Minimum Deviance:", min(cv_model$cvm), "\n")
}

cv_en <- cv.glmnet(X_train, y_train, family = "binomial", alpha = 0.2, standardize = TRUE)

plot(cv_en)

en_coefs <- coef(cv_en, s = "lambda.1se")
print(en_coefs)
```

From the results, the optimal value is 0.2 since it has the lowest minimum deviance. The elastic net regularization is then run, which reduces the model to 17 variables. logistic regression model is created below using the 17 variables to start:

```{r log_regress, warning = FALSE}
fit_glm_cv <- glm(Default ~ Limit_Bal+Marriage+Age+Pay_Sep+Pay_Aug+Pay_July+
                  Pay_June+Bill_Amt_Sep+Pay_Amt_Sep+Pay_Amt_Aug+
                  Pay_Amt_July+Pay_Amt_June+Pay_Amt_May,
                  data = train_set, family = binomial())

summary(fit_glm_cv)
stepAIC(fit_glm_cv)
#VIF of the 2nd logistic regression model. VIF scores appear good.
vif(fit_glm_cv)

fit_glm_cv2 <- glm(Default ~ Limit_Bal+Marriage+Age+Pay_Sep+Pay_Aug+Pay_July+
                  Pay_June+Bill_Amt_Sep+Bill_Amt_June+Pay_Amt_Sep+Pay_Amt_Aug+Pay_Amt_June,
                  data = train_set, family = binomial())

summary(fit_glm_cv2)
stepAIC(fit_glm_cv2)
#VIF of the 2nd logistic regression model. VIF scores appear good.
vif(fit_glm_cv2)
#Elastic net logistic regression model - removed bill_amt_aug (high VIF)
#Removed pay_june and pay_may since they were not statistically significiant
fit_glm_cv3 <- glm(Default ~ Limit_Bal+Education+Marriage+Age+Pay_Sep+Pay_Aug+Pay_July+
                  Pay_May+Bill_Amt_Sep+Pay_Amt_Sep+Pay_Amt_Aug
                  +Pay_Amt_July+Pay_Amt_June,
                  data = train_set, family = binomial())

summary(fit_glm_cv3)
stepAIC(fit_glm_cv3)
#VIF of the 3rd logistic regression model. VIF scores appear good.
vif(fit_glm_cv3)


```


Four variables are discarded due to either statistical insignificance in the model or a high VIF value, indicating high multicollinearity between multiple variables, leading to 13 variables being kept.

```{r predictions, warning=FALSE}
pred_probs <- predict.glm(fit_glm_cv3, newdata = test_set, type = "response")
#Displays the predictions for a few values. 
head(pred_probs)
#Sorts predictions into their respective class (0 or 1) depending on their value.
pred <- ifelse(pred_probs<0.5, 0,1)
#Creates and displays the confusion matrix table based on the actual and predicted values.
confusion_table <- table(test_set$Default, pred)
confusion_table
#Creates the confusion matrix statistics for the logistic regression model.
cm_log <- confusionMatrix(confusion_table, positive = '1', mode = "everything")
#Saves the accuracy, precision, and recall values.
log_accuracy = accuracy(test_set$Default, pred)
log_precision = cm_log$byClass['Precision']
log_recall = cm_log$byClass['Recall']
log_pos_precision = cm_log$byClass['Neg Pred Value']
#Prints the accuracy, precision, and recall values.
print(paste("Accuracy: ", round(log_accuracy,3)))
print(paste("Precision: ", round(log_precision,3)))
print(paste("Recall: ", round(log_recall,3)))
print(paste("Default Precision: ", round(log_pos_precision,3)))
```
Once the model was run, the accuracy, precision, and recall were found for the prediction model. Accuracy describes how often the model is correct in its overall prediction. Precision identifies how often the model identifies those who default on their credit card out of all who do so, while recall identifies how often the model correctly identifies those who default on their credit card.Another way of describing precision and recall is precision is a measure of quality, while recall is a measure of quantity.

In the case of the logistic regression model, the accuracy was `r round(log_accuracy,3)`\%, precision was `r round(log_precision,3)`\%, and recall was `r round(log_recall,3)`\%. The precision for predicting actual default cases correctly was `r round(log_pos_precision,3)`\%. Overall, the logistic regression model performed well for predicting whether a client would default even though the model contained a lot of false positives by predicting a client would default when in reality they would not. In the case of credit risk, it is best to have a higher recall rate compared to precision since it is better to have false positives occur compared to false negatives as they cost less money for the company. A false positive leads to higher operational costs for the company (sending payment reminders, lowering credit limit if applicable) while a false negative could cost the company thousands more if a client actually defaults on their payment. 

The regression model's accuracy, specificity, and sensitivity can be improved by optimizing the cutoff point for the model. One way of doing so is using the Receiver Operating Characteristic (ROC) curve, which plots the true positive (sensitivity) against the false positive rate against various thresholds. The AUC curve can be used to measure the performance of the model, with a higher AUC number demonstrating better model performance. An AUC 0.8 and above indicates good model performance. The model has an AUC score of 0.719, which indicates acceptable model performance. This will be a factor to keep in mind for potential model improvements. 

```{r opt_cuts, warning=FALSE}
prob <- predict(fit_glm_cv3, type = "response")
# Create ROC curve
roc_obj <- roc(train_set$Default, prob, plot = TRUE, col = "blue", print.auc = TRUE)

# Find optimal cutoff (maximizes sensitivity + specificity)
#AUC performance = 0.718
opt <- coords(roc_obj, "best", ret = c("threshold", "sensitivity", "specificity"))
print(opt)
#threshold = 0.251
#sensitivity = 0.548
#specificity = 0.822
```

The ROC curve supplies various cutoff values for the logistic regression model. From the ROC curve, the optimal cutoff point to maximize all three measurements (accuracy, specificity, and sensitivity) is 0.251, while the cutoffs to maximize sensitivity and specificity are 0.548 and 0.822, respectively. The optimal cutoff from the AUC curve, which maximizes both sensitivity and specificity, is 0.719. There is a tradeoff for each cutoff point, so it is up to the user to determine which is best for the purpose of the model. I chose to use 0.548 since I want to maximize both the overall accuracy of the model while accurately identifying those who will default on their credit card. 

```{r cutoff_preds, warning=FALSE}
pred_new <- ifelse(pred_probs<0.529, 0,1)
#Creates and displays the confusion matrix table based on the actual and predicted values.
confusion_table_new <- table(test_set$Default, pred_new)
confusion_table_new
#Creates the confusion matrix statistics for the logistic regression model.
cm_log_opt <- confusionMatrix(confusion_table_new, positive = '1', mode = "everything")
#Saves the accuracy, precision, and recall values.
log_accuracy_opt = accuracy(test_set$Default, pred_new)
log_precision_opt = cm_log_opt$byClass['Precision']
log_recall_opt = cm_log_opt$byClass['Recall']
log_pos_precision_opt = cm_log_opt$byClass['Neg Pred Value']
#Prints the accuracy, precision, and recall values.
print(paste("Accuracy: ", round(log_accuracy_opt,3)))
print(paste("Precision: ", round(log_precision_opt,3)))
print(paste("Recall: ", round(log_recall_opt,3)))
print(paste("Default Precision: ", round(log_pos_precision_opt,3)))
```

From the results, the recall, and default precision improved with the new threshold value, while accuracy and precision had decreased. Using this new threshold helps predict a default more accurately, but does so by incorrectly labeling someone as 'default' when they would not do so more often than the previous threshold mark. Overall, this model is kept since the overall accuracy improves and it's more important to identify those who will default. 

## Random Forest

Another prediction model used is random forest. Random forest is a classifying method consisting of many decision trees. By creating a "forest" of decision trees, the classifying model hopes to select it's best model by running many different decision trees and "takes the majority" to determine classification. To do so, random forest uses out-of-bag sampling.

A random forest model is created to determine the probability of credit card default:

```{r forest_normal}
set.seed(123)

n_defaults <- sum(train_set$Default == 1)

#Random Forest for variables. mtry = 4 since there are 13 variables (square root of 11 is close to 4).
fit_rf <- randomForest(factor(Default) ~Limit_Bal+Education+Marriage+Age+Pay_Sep+Pay_Aug+Pay_July+
                  Pay_May+Bill_Amt_Sep+Pay_Amt_Sep+Pay_Amt_Aug
                  +Pay_Amt_July+Pay_Amt_June, 
                  mtry = 4, ntree = 1000, sampsize = c(n_defaults, n_defaults),
                   strata = factor(train_set$Default), data = train_set)
varImpPlot(fit_rf)
#Predicts values in the test set.
predict_rf <- predict(fit_rf, test_set)
#Creates the confusion matrix table for the random forest model.
confusion_table_rf <- table(test_set$Default, predict_rf)
#Creates and displays the confusion matrix statistics for the random forest model.
cm_rf <- confusionMatrix(confusion_table_rf, positive = '1', mode = "everything")
cm_rf
#Saves the accuracy, precision, and recall values.
rf_accuracy = accuracy(test_set$Default, predict_rf)
rf_precision = cm_rf$byClass['Precision']
rf_recall = cm_rf$byClass['Recall']
rf_pos_precision = cm_rf$byClass['Pos Pred Value']
#Prints the accuracy, total precision, recall, and default precision values.
print(paste("Accuracy: ", round(rf_accuracy,3)))
print(paste("Precision: ", round(rf_precision,3)))
print(paste("Recall: ", round(rf_recall,3)))
print(paste("Default Precision: ", round(rf_pos_precision,3)))
```

From the input printed and the plot provided, it is seen that the pay September variable and bill amount in September, as well as age are important variables in determining credit card default. It can also be argued pay amount in September, pay amount in August and pay amount in July are important variables in determining credit card default. 

Looking at the confusion matrix, the random forest model's accuracy was `r round(rf_accuracy,3)`\%, precision was `r round(rf_precision,3)`\%, and recall was `r round(rf_recall,3)`\%. The precision for predicting actual default cases correctly was `r round(rf_pos_precision,3)`\%. Overall, the random forest model was slightly better in its accuracy and much better for its precision. However, the logistic regression model performed better than the random forest model when predicting actual default cases, as evidenced by the recall and default precision values. 

## Conclusion

In this project, logistic regression and random forest models were created to predict if an individual would default on their credit card.

First, the data was cleaned for accuracy and manipulated to view distributions and trends in the data. From the tables and plots created, the data had more females than males and was skewed in age, with participants below the age of 40 much more prevalent than participants over the age of 40.

Next, prediction models were created to predict an individual's chances of defaulting on their credit card. The first model used was a logistic regression model. This model was used to predict if an individual would default on their credit card based on their information. This model is great for predicting a Yes/No classification for individuals. From the model created, three individuals were created with their unique information. In the example above, all three individuals created had a good chance of not defaulting on their credit card.

A random forest model was also created to determine the most important variables in a prediction model, as well as to see the accuracy of the created model. From the results, the random forest model could accurately predict someone not defaulting on their credit card, but had a more difficult time accurately predicting when someone would default on their credit card.

When comparing the two models, the following table was created:

```{r pander}
set.caption("Performance for Logistic Regression and Random Forest Models")
data.table = rbind(c(log_accuracy_opt, log_precision_opt, log_recall_opt, log_pos_precision_opt), c(rf_accuracy, rf_precision, rf_recall, rf_pos_precision))
colnames(data.table) = c("Accuracy", "Precision", "Recall", "Default Precision")
rownames(data.table) = c("Logistic Regression", "Random Forest")

pander(data.table)
```

Overall, it seems the logistic regression model has a higher recall and default precision rate than the random forest model, but does worse than the random forest model in accuracy and precision. This means the logistic regression model has less false negatives than the random forest model, but also has more false positives. Though the accuracy seems fairly high for the random forest prediction model, I am concerned with the false negative rates and the low default precision percentage.

Though these prediction models are acceptable, there is room for improvement, particularly in accurately predicting client that will default. I believe adding certain variables such as credit score, credit age, and credit card utilization can help improve the prediction models. 

Thank you for viewing my project.

# END
